Abstract Data Type Inference

### 2015-03-24 histories.tex
Formula example.

%!TEX root = draft.tex
\section{Histories of ADT Implementations}
\label{sec:histories}

TODO MOTIVATE THIS SECTION

We fix an arbitrary infinite set $\mathbb{O}$ of operation identifiers, and
given sets $\mathbb{M}$ and $\mathbb{V}$ of method names and values. A
\emph{call action} binds an operation $o \in \mathbb{O}$ with a method name $p
\in \mathbb{M}$ and argument value $v \in \mathbb{V}$, while a \emph{return
action} binds an operation $o \in \mathbb{O}$ with a return value $v \in
\mathbb{V}$. An \emph{execution} $e$ is a sequence of call and return actions
where
\begin{itemize}

@@

\begin{example}

>>>>

\section{Histories of ADT Implementations}
\label{sec:histories}

Implementations of abstract data types (ADTs) provide methods which are called,
possibly concurrently, by client programs. We capture the possible logs of
method calls and returns in executions of client programs abstractly, in order
to facilitate the development of symbolic ADT representations.

To this end, we fix an arbitrary infinite set $\mathbb{O}$ of operation
identifiers, and given sets $\mathbb{M}$ and $\mathbb{V}$ of method names and
values. A \emph{call action} binds an operation $o \in \mathbb{O}$ with a
method name $p \in \mathbb{M}$ and argument value $v \in \mathbb{V}$, while a
\emph{return action} binds an operation $o \in \mathbb{O}$ with a return value
$v \in \mathbb{V}$. An \emph{execution} $e$ is a sequence of call and return
actions where
\begin{itemize}

@@

\begin{example}
  \label{ex:histories-of-executions}

### 2015-03-24 inference.tex

TODO MOTIVATE THIS SECTION

An \emph{abstract data type (ADT)} $A$ is the set $H(\mathcal{I})$ of histories
of an implementation $\mathcal{I}$. A \emph{kernel} of an ADT $A$ is a minimal
set $H$ of histories generating $A$, i.e.,~such that
\begin{align*}
  A = \set{ h' : \exists h \in H.\ h \to^\ast h' }
\end{align*}
where $\to = (\to_\mathrm{o} \cup \to_\mathrm{p} \cup \to_\mathrm{c})$. In
Section~\ref{sec:nature} we demonstrate that naturally-occurring ADTs have
unique, bounded-width kernels. For the remainder of this work, we assume ADTs
have unique, bounded-width kernels, denoting \emph{the kernel} of $A$ by
$\ker A$. When $\ker A$ has width $1$, we say $A$ is \emph{atomic}.

@@

Recent works demonstrate efficient algorithms\footnote{In time polynomial in
the number of operations, per execution.} for checking observational
refinement~\cite{conf/popl/BouajjaniEEH15, conf/pldi/EmmiEH15}, yet rely on
hand-written symbolic ADT representations. In order to frame the problem of
computing these automatically, we fix a language for symbolic representation. A
\emph{history formula} $F$ is a first-order logic formula with
\begin{itemize}

@@

  The following history formula is satisfied by histories in which no write
  operation happens between a pair of matching write and read operations:
  \begin{align*}
    \forall x, y, z.\ 
    & {\sf c}(x) \land {\sf f}(x) = \mathrm{write} \land {\sf m}(x) = x \\
    & \land {\sf c}(y) \land {\sf f}(y) = \mathrm{write} \land {\sf m}(y) = y \\
    & \land {\sf c}(z) \land {\sf f}(z) = \mathrm{read} \land {\sf m}(z) = x \\
    & \land x < y 
      \implies z < y
  \end{align*}
  This is one of several requirements of atomic single-value register ADTs,
  which is satisfied by certain linearizations of histories $H(e_2)$ and
  $H(e_3)$ of Example~\ref{ex:histories-of-executions}, yet not $H(e_1)$.

@@

  The \emph{symbolic ADT inference problem} is to compute a history formula $F$
  representing the ADT $H(\mathcal{I})$ of a given implementation $\mathcal{I}$.

>>>>

In this section we formalize a notion of abstract data type and define the
corresponding refinement and inference problems. These are the foundational
problems addressed in this work.

An \emph{abstract data type (ADT)} $A$ is the set $H(\mathcal{I})$ of histories
of an implementation $\mathcal{I}$. A \emph{kernel} of an ADT $A$ is a minimal
set $H$ of histories generating $A$, i.e.,~such that
\begin{align*}
  A = \set{ h' : \exists h \in H.\ h \to^\ast h' }
\end{align*}
where $\mathord\to = (\to_\mathrm{o} \cup \to_\mathrm{p} \cup \to_\mathrm{c})$.
In Section~\ref{sec:nature} we demonstrate that naturally-occurring ADTs have
unique, bounded-width kernels. For the remainder of this work, we assume ADTs
have unique, bounded-width kernels, denoting \emph{the kernel} of $A$ by $\ker
A$. When $\ker A$ has width $1$, we say $A$ is \emph{atomic}.

@@

Recent works demonstrate efficient algorithms\footnote{In time polynomial in
the number of operations, per execution.} for checking observational
refinement~\cite{conf/popl/BouajjaniEEH15, conf/pldi/EmmiEH15}, yet rely on
hand-written symbolic ADT representations. In order to frame the problem of
computing these automatically, we fix a language for symbolic representation. A
\emph{history formula} is a first-order logic formula with
\begin{itemize}

@@

  The following history formula is satisfied by histories in which no write
  operation happens between a pair of matching write and read operations:
  \begin{align*}
    \forall x_1, x_2, x_3.\ 
    & {\sf c}(x_1) \land {\sf f}(x_1) = \mathrm{write} \land {\sf m}(x_1) = x_1 \\
    & \land {\sf c}(x_2) \land {\sf f}(x_2) = \mathrm{write} \land {\sf m}(x_2) = x_2 \\
    & \land {\sf c}(x_3) \land {\sf f}(x_3) = \mathrm{read} \land {\sf m}(x_3) = x_1 \\
    & \land x_1 < x_2 
      \implies x_3 < x_2
  \end{align*}
  This is one of several requirements of atomic single-value register ADTs,
  which is satisfied by certain linearizations of histories $H(e_2)$ and
  $H(e_3)$ of Example~\ref{ex:histories-of-executions}, yet not $H(e_1)$.

@@

  The \emph{symbolic ADT inference problem} is to compute a history formula
  representing the ADT $H(\mathcal{I})$ of a given implementation $\mathcal{I}$.

### 2015-03-24 nature.tex

TODO THERE ARE 24 PATTERNS FOR EACH OF THE 4 VARIATIONS

>>>>

The work-stealing queue~\cite{journals/dc/HendlerLMS06} implements a collection
of data values with first-in-first-out\footnote{There are actually four
variations to the work-stealing queue, depending on the ends from which the
take and steal operations remove values. While our approach works
indifferently, below we focus on the variation where both remove the least
recent.} (FIFO) removal order, proving three methods:
\begin{itemize}

  \item give$(v)$ adds the value $v$ to the queue,

  \item take $\Rightarrow v$ removes the value $v$, and

  \item steal $\Rightarrow v$ removes the value $v$.

\end{itemize}
Unlike atomic queues, the work-stealing queue permits values to be removed
twice: once normally, via the take operation, and once exceptionally, via the
steal operation. We give a faithful matching scheme over executions which add
unique values analogously to the scheme for stacks and queues, in which give
operations match themselves, while take and steal operations match the give
operation which added their returned value, or themselves, in case the nil
value is returned. Note that the kernel of work-stealing queues has width $2$,
since a pairs of concurrent take and steal operations returning the same value
are permitted, while sequentially they are not. As our implementation is
currently limited to width-$1$ histories (see the discussion in
Section~\ref{sec:discussion}), below we generate only the width-$1$ complement.

We compute the following twenty-four histories to generate the complement of
the work-stealing queue ADT:
\begin{verbatim}
  [1:_] take => 1  #             [1:_] steal => 1  #
  ---                            ---
  [1:1] give(1)         #        [1:1] give(1)          #
  [2:2] take => - (RO)    #      [2:2] steal => - (RO)    #
  ---                            ---
  [1:2] take => 1  #             [1:2] steal => 1  #
  [2:2] give(1)      #           [2:2] give(1)       #
  ---                            ---
  [1:1] give(1)    #             [1:1] give(1)     #
  [2:2] give(2)      #           [2:2] give(2)       #
  [3:2] take => 2      #         [3:2] steal => 2      #
  ---                            ---
  [1:1] give(1)    #             [1:1] give(1)     #
  [2:1] take => 1    #           [2:1] take => 1     #
  [3:1] take => 1      #         [3:1] steal => 1      #
  ---                            ---
  [1:1] give(1)         #        [1:1] give(1)         #
  [2:2] take => - (RO)    #      [2:2] take => - (RO)    #
  [3:1] take => 1           #    [3:1] steal => 1          #
  ---                            ---
  [1:1] give(1)     #            [1:1] give(1)     #
  [2:1] steal => 1    #          [2:1] steal => 1    #
  [3:1] take => 1       #        [3:1] steal => 1      #
  ---                            ---
  [1:1] give(1)          #       [1:1] give(1)          #
  [2:2] steal => - (RO)    #     [2:2] steal => - (RO)    #
  [3:1] take => 1            #   [3:1] steal => 1           #
  ---                            ---
  [1:2] take => 1   #            [1:3] take => 1   #
  [2:2] give(1)       #          [2:3] steal => 1    #
  [3:2] steal => 1      #        [3:3] give(1)         #
  ---                            ---
  [1:2] steal => 1  #            [1:3] steal => 1  #
  [2:2] give(1)       #          [2:3] take => 1     #
  [3:2] take => 1       #        [3:3] give(1)         #
  ---                            ---
  [1:1] give(1)    #             [1:1] give(1)     #
  [2:2] give(2)      #           [2:2] give(2)       #
  [3:2] take => 2      #         [3:2] take => 2       #
  [4:1] take => 1        #       [4:1] steal => 1        #
  ---                            ---
  [1:1] give(1)     #            [1:1] give(1)     #
  [2:2] give(2)       #          [2:2] give(2)       #
  [3:2] steal => 2      #        [3:2] steal => 2      #
  [4:1] take => 1         #      [4:1] steal => 1        #
\end{verbatim}
As new generators are discovered for $n=1$ and $n=2$ matches only, the
work-stealing queue ADT is predictable.
